---
title: | 
      | Base R
      | 
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: zenburn
    css: documentCSS.css
---
```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE, comment = "")
```

# Symbols

Here are some basic symbols that you will see in R:

## <-

An assignment operator
  
```{r}
x <- 1:5
```

  
## =   
  
An alternative assignment operator (not preferred by the purists)
  
```{r}
y = 6:10
```
  
## ()  
  
() will always be for a function -- for example mean() 

```{r}
mean(x)
```
  
## \{\}  
  
{} always contain calls to created functions

```{r}
function(x) {
  x + 5
}
```

## \[\]
  
[] are used for indexing

```{r}
# Gets the second thing
x[2] 

# Gets the first row
cars[1, ] 

# Gets the first column
cars[, 1] 

# Gets the value in the first row
# in the first column
cars[1, 1] 


# Gets the first 10 rows of the "dist" column
cars[1:10, "dist"]
```  


## $  

$ is a shortcut used for accessing names within a data.frame

```{r}
cars$speed
```

It is practically equivalent to:

```{r}
cars[, "speed"]
```

## "" and ''
  
"" and '' are both used to specify strings

```{r}
z <- c("string1", 'string2', "string3")
```

## ``
  
`` are used for making something literal

```{r}
cars$`1 $ bad name` = 1:nrow(cars)

head(cars)

`$`(cars, "speed")
```

## ==
  
== is strictly equivalent

```{r}
cars[cars$speed == 20, ]
```

## !=
  
!= is not equal to

```{r}
cars[cars$speed != 20, ]
```

## |  

| is an or operator

```{r}
cars[cars$speed == 18 | cars$speed == 20, ]
```

## &

& means and

```{r}
cars[cars$speed == 18 & cars$dist == 56, ]
```

## %*%
  
%*% is matrix multiplication

```{r}
x %*% y
```

## %%
  
%% is the modulus

```{r}
10 %% 4
```

## |> 

|> is a pipe that passes the output of one function into another function:

```{r}
1:10 |> 
  mean()
```

See the magrittr package for more potential pipes.

# Finding Packages

<a href="https://cran.r-project.org/web/views/">CRAN Task Views</a> will help you find packages that do what you want to do. They are arranged into broad categories that then get broken down.

There is also <a href="https://rseek.org/">rseek</a>

# Package Installation

Install new packages from CRAN with:

```{r, eval = FALSE}
install.packages("tidyverse")
```

If you want to install multiple, you need to use a character vector:

```{r, eval = FALSE}
install.packages(c("tidyverse", "psych", "Hmisc"))
```

If you want to install something from Github:

```{r, eval = FALSE}
devtools::install_github("tidyverse/ggplot2")
remotes::install_github("tidyverse/ggplot2")
```

# Loading Packages

If you want to use a package, you load it into your local environment with:

```{r, eval = FALSE}
library(ggplot2)
```

# Data Types

There are a few different types of data within R.

We have numeric variables:

```{r}
nums <- 1:10

nums
```

We have characters:

```{r}
chars <- c("Poor", "Fair", "Good", "Great")

chars
```

And we have factors:

```{r}
facs <- as.factor(chars)

facs
```

We can have just regular factors like above or we can order those factors:

```{r}
orderedFacs <- ordered(facs, 
                      levels = c("Poor", "Fair", "Good", "Great"))

orderedFacs
```

And logicals:

```{r}
logicalVector <- c(TRUE, FALSE, TRUE, FALSE, FALSE)

sum(logicalVector)

logicalVector[1] == 1
```


# Data Structures

At our most basic, we have a vector. Think of it a just a basic list of numbers:

```{r}
vectorExample <- rnorm(10)

vectorExample
```

There is the matrix. A matrix is a rectangle where everything is the same type of data:

```{r}
characterMatrix <- matrix(letters, nrow = 5, ncol = 5)

characterMatrix
```

```{r}
numericMatrix <- matrix(1:25, nrow = 5, ncol = 5)

numericMatrix
```

If we have data of different types, we are dealing with data frames:

```{r}
dfExample <- data.frame(numVar = 1:3, 
                        charVar = letters[1:3], 
                        facVar = factor(c("Poor", "Good", "Great")), 
                        ordVar = ordered(c("Poor", "Good", "Great"), 
                                         levels = c("Poor", "Good", "Great")))

summary(dfExample)
```

```{r}
str(dfExample)

is.data.frame(dfExample)

is.list(dfExample)
```

We also have lists. Lists can contain any number of any thing within the different list entries:

```{r}
listExample <- list(1, 
                    rnorm(10), 
                    numericMatrix, 
                    rnorm(30), 
                    list(dfExample, cars))

listExample
```

Whenever you have lists, we have an extra set of indices to consider:

```{r}
listExample[[3]]
```

```{r}
listExample[[5]][[2]]
```

As play with more data (especially from the web), you will see even more lists.

# Useful Functions

There are many functions that will be helpful, but here are a few easy ones:

```{r}
sum(c(1:10, NA), na.rm = TRUE)

mean(1:10, na.rm = TRUE)

sd(1:10, na.rm = TRUE)

paste("tic", "tac", "toe", sep = "-")

state.abb

match(c("January", "February"), month.name)

testWords <- c("Bad", "Poor", "Great", "Awesome")

dfExample[dfExample$facVar %in% testWords, ]

dfExample[!(dfExample$facVar %in% testWords), ]
```

There are also a lot of functions that ask specific questions:
  
```{r}
set.seed(1001)

test_df <- data.frame(x = rnorm(100), 
                      y = runif(100, 1, 5))

which.max(test_df$x)

test_df[which.max(test_df$x), ]

any(test_df$x > 3)

all(test_df$x < 3)

test_df[which(test_df$x > 3), ]

test_df$max_val <- pmax(test_df$x, test_df$y)

anyNA(test_df$x)

is.na(test_df)
```

There are also a huge number of ways to describe data:

```{r}
data("msleep")

msleep <- as.data.frame(msleep)

table(msleep$vore)

prop.table(table(msleep$vore))

prop.table(table(msleep$vore, msleep$conservation))

prop.table(table(msleep$vore, msleep$conservation), margin = 1)

prop.table(table(msleep$vore, msleep$conservation), margin = 2)

addmargins(table(msleep$vore, msleep$conservation))
```


# Selecting 

## Base

There are many ways to select variables with <span class="pack">base</span> R:

```{r, eval = FALSE}
msleep[, c(1:5, 7:8)]

keepers <- c("name", "genus", "vore", "sleep_total", "sleep_rem")

msleep[, keepers]

msleep[, c("name", grep("^sleep", names(msleep), values = TRUE))]
```

You can also drop variables:

```{r}
msleep[, -c(1:2)]

dropVars <- c("conservation", "vore")

msleep[, !(names(msleep) %in% dropVars)]
```


```{r}
library(lavaan)

testData <- HolzingerSwineford1939

names(testData)

keepers <- c(grep("^age", names(testData), value = TRUE), 
            paste("x", 1:9, sep = ""))

testData <- testData[, keepers]
```


# Subsetting/Filtering 

One of the more frequent tasks is related to filtering/subsetting your data. You often want to impose some type of rules on your data (e.g., US only, date ranges).

R gives us all the ability in the world to filter data.

```{r}
summary(msleep[msleep$sleep_total < mean(msleep$sleep_total), ])
```

# New Variables and Recoding 

Adding a new variable in <span class="pack">base</span> R is as easy as the following:

```{r}
msleep$roundedSleep <- round(msleep$sleep_total)
```

You will need to recode variables at some point. Depending on the nature of the recode it can be easy (e.g., to reverse code a scale, you just subtract every value by max value + 1).

You will need to do some more elaborate stuff:

```{r, eval = FALSE}
msleep$sleepLoHi <- 0

msleep$sleepLoHi[msleep$sleep_total > median(msleep$sleep_total)] <- 1
```

```{r, eval = FALSE}
msleep$mpgLoHi <- ifelse(msleep$sleep_total > median(msleep$sleep_total), 1, 0)
```

# The apply family

In addition to for loops, R has a whole set of functions for applying functions over various parts of your data. Let's start with the most general type: apply.

We can use apply like this:

```{r}
apply(X = msleep[, c("sleep_total", "sleep_rem", "sleep_cycle")], 
      MARGIN = 1, 
      FUN = function(x) mean(x, na.rm = TRUE))
```

What we just did was to apply the mean function over the rows (MARGIN = 1) of the cars data frame. We could even assign that to our data:

```{r}
msleep$sleepMeans <- apply(X = msleep[, c("sleep_total", "sleep_rem", "sleep_cycle")], 
      MARGIN = 1, 
      FUN = function(x) mean(x, na.rm = TRUE))
```

If `MARGIN = 1` works over the rows, what would be the alternative:

```{r}
apply(X = msleep[, c("sleep_total", "sleep_rem", "sleep_cycle")], 
      MARGIN = 2, 
      FUN = function(x) mean(x, na.rm = TRUE))
```

Whenever you need to apply a function over the rows or columns of a data frame, apply has you covered.

One of the most useful members of the apply family is lapply. You can give it a vector or a list, perform a task over that object, and then always get a list returned.

```{r}
test_values <- list(std_normal = rnorm(10000), 
                    iq_norm = rnorm(1000, 100, 15), 
                    score_norm = rnorm(100, 27, 3))

lapply_result <- lapply(test_values, mean)

data.frame(lapply_result)
```

Perhaps the most interesting thing that lapply can do is to work over a data frame, since a data frame is nothing more than a fancy list.

```{r}
str(msleep)

factor_vars <- c("genus", "vore", "order", "conservation")

msleep[, factor_vars] <- lapply(factor_vars, 
                                function(x) as.factor(msleep[, x]))

str(msleep)
```

Finally, we come to mapply: multiple apply. The goal of mapply is to apply several different vectors over a function:

```{r}
mapply(mean, list(1:10, 20:30, c(20, 340, NA)), na.rm = TRUE)
```

Let's compare what that would look like with an lapply:

```{r}
lapply(list(1:10, 20:30, c(20, 340, NA)), mean, na.rm = TRUE)
```

We can see that lapply returns a list, whereas mapply returns a simplified vector!

The real power of mapply comes in when you need to create your own function:

```{r}
my_func <- function(x, y) {
  
  result <- x * y
  
  data.frame(x = x, 
             y = y, 
             result = result)
}

results <- mapply(my_func, x = 1:10, y = 1:10, SIMPLIFY = FALSE)

do.call('rbind', results)
```

While that isn't anything that we couldn't do in a data frame already, those values will recycle if there is an unequal length:

```{r}
results <- mapply(my_func, x = 1:10, y = 1:20, SIMPLIFY = FALSE)

do.call('rbind', results)
```

The mapply function also provides an excellent alternative to most nested for loops:

```{r}
nba_data <- read.csv("https://www.nd.edu/~sberry5/data/nba_data.csv")

# players <- unique(nba_data$PLAYER_ID)

teams <- unique(nba_data$TEAM_ID)

results_list <- list(list())

for(i in teams) {
  team_name <- unique(nba_data$TEAM_NAME[nba_data$TEAM_ID == i])
  
  players <- unique(nba_data$PLAYER_ID[nba_data$TEAM_ID == i])
  
  for(j in players) {
    player_points <- sum(nba_data$PTS[nba_data$PLAYER_ID == j])
    player_name <- unique(nba_data$PLAYER_NAME[nba_data$PLAYER_ID == j])
    result <- data.frame(player_points, player_name, team_name)
    results_list[[length(results_list) + 1]] <- result
  }
}

results_list <- do.call('rbind', results_list)
```

That beast becomes something as simple as this:

```{r}
players_teams <- nba_data[, c("PLAYER_ID", "TEAM_ID")]

players_teams <- unique(players_teams)

point_function <- function(player, team) {
  team_name <- unique(nba_data$TEAM_NAME[nba_data$TEAM_ID == team])
  
  player_points <- sum(nba_data$PTS[nba_data$PLAYER_ID == player & 
                                      nba_data$TEAM_ID == team])
  
  player_name <- unique(nba_data$PLAYER_NAME[nba_data$PLAYER_ID == player])
  
  result <- data.frame(player_points, player_name, team_name)
  
  return(result)
}

result <- mapply(point_function, 
       players_teams$PLAYER_ID, 
       players_teams$TEAM_ID, SIMPLIFY = FALSE)

result <- do.call('rbind', result)
```

In the end, that particular function is a long way around doing an aggregate/group_by & summarize.

### Iteration

Before moving on from this class, the concept of iteration needs to be crystal clear and we have just seen several versions of it. 

Let's imagine that we have a set of variables; we can just use the variables from msleep 

For this instance let's say that we want to try every variable in that data as an individual predictor; please note, though, that this is something that you would **never** do in practice. There are better methods for handling this procedure. 

Let's see how we would structure this in a for loop:

```{r}
predictor_vars <- msleep[, sapply(colnames(msleep), function(x) is.numeric(msleep[, x]))] |>
  colnames()

predictor_vars <- predictor_vars[-which(predictor_vars == "brainwt")] 

predictor_count <- length(predictor_vars)

results_df <- data.frame(predictor = character(predictor_count), 
                         coefficient = numeric(predictor_count), 
                         rsquared = numeric(predictor_count))

for(i in 1:predictor_count) {
  predictor <- predictor_vars[i]
  
  model_formula <- as.formula(paste0("brainwt ~ ", predictor))
  
  regression_model <- lm(model_formula, data = msleep)
  
  model_result <- data.frame(predictor = predictor, 
                             coefficient = regression_model$coefficients[predictor], 
                             rsquared = model_summary$r.squared)
  
  model_summary <- summary(regression_model)
  
  results_df[i, ] <- model_result
}

results_df
```

We can do the exact same thing with an lapply:

```{r}
model_results <- lapply(predictor_vars, function(predictor) {
  
  model_formula <- as.formula(paste0("brainwt ~ ", predictor))
  
  regression_model <- lm(model_formula, data = msleep)
  
  model_summary <- summary(regression_model)
  
  model_result <- data.frame(predictor = predictor, 
                             coefficient = regression_model$coefficients[predictor], 
                             rsquared = model_summary$r.squared)
  
  return(model_result)
})

model_results <- do.call(rbind, model_results)
```


Or with map_df:

```{r}
model_results <- purrr::map_df(.x = predictor_vars, ~{
  model_formula <- as.formula(paste0("brainwt ~ ", .x))
  
  regression_model <- lm(model_formula, data = msleep)
  
  model_summary <- summary(regression_model)
  
  model_result <- data.frame(predictor = .x, 
                             coefficient = regression_model$coefficients[.x], 
                             rsquared = model_summary$r.squared)
  
  return(model_result)
})
```

Please notice that there is absolutely zero difference in what you get out of these functions. The only difference is the work that you need to do to make it work.

# Data Import 

Importing data is often the easiest part (never too hard to import a nice .csv). Sometimes, though, we need some other strategies.

## Delimited Files

Frequently, you will see nicely delimited text files that are not .csv files -- these are often tab-delimited files, but they can take other forms. 

```{r, eval = FALSE}
read.table("https://download.bls.gov/pub/time.series/ce/ce.data.42a.RetailTrade.Employment", 
           header = TRUE, sep = "\t")
```

Is the same as:

```{r, eval = FALSE}
read.delim("https://download.bls.gov/pub/time.series/ce/ce.data.42a.RetailTrade.Employment")
```

The <span class="func">read.table()</span> function gives you added flexibility to specify many different parameters. 

Examine the following file from SDC Platinum and read it in properly:

<a href="https://www3.nd.edu/~sberry5/data/sdcTest.txt">SDC Wackiness</a>

How did you do?

Did you notice anything about these files? They are not really very big, but they might have taken a little bit of time to read in. There have been times where people have commented that R is too slow on the read side. If you find you files are not being read quickly enough, you can try a few alternatives: <span class="pack">readr</span> and <span class="pack">data.table</span>

Try the following:

```{r, eval = FALSE}
library(readr)

readrTest <- read_delim("https://download.bls.gov/pub/time.series/ce/ce.data.42a.RetailTrade.Employment", 
                       delim = "\t")
```


```{r, eval = FALSE}
library(data.table)

dtTest <- fread("https://download.bls.gov/pub/time.series/ce/ce.data.42a.RetailTrade.Employment", 
               sep = "\t")
```

That SDC file that might have taken a few minutes will now take just a few seconds:

```{r, eval = FALSE}
sdc <- read_delim("https://www3.nd.edu/~sberry5/data/sdcTest.txt", 
                 delim = "^")
```

Pretty awesome, right?

At times, you will get data in some proprietary format. That is when you need to turn to other places.

## Excel

Download the following Excel file: https://www3.nd.edu/~sberry5/data/excelTest.xlsx

```{r, eval = FALSE}
readxl::read_excel(path = "")
```

What do we know about Excel workbooks? Check out the help on <span class="pack">readxl</span> and let me know our path forward.

## SAS

```{r, eval = FALSE}
haven::read_sas(data_file = "https://www3.nd.edu/~sberry5/data/wciklink_gvkey.sas7bdat")
```

## Stata

```{r, eval = FALSE}
haven::read_dta(file = "https://www3.nd.edu/~sberry5/data/stataExample.dta")
```

## SPSS

We often see the -99 added as the missing value in SPSS (of course, there is no way that -99 would ever be an actual value, right?).

```{r, eval = FALSE}
haven::read_spss(file = "https://www3.nd.edu/~sberry5/data/spssExample.sav", 
                 user_na = "-99")
```

## HTML

Depending on your needs, reading an html table into R is getting to be too easy.

```{r, eval = FALSE}
library(rvest)

cpi <- read_html("http://www.usinflationcalculator.com/inflation/consumer-price-index-and-annual-percent-changes-from-1913-to-2008/") %>% 
  html_table()
```

Things might get a bit tricky:

```{r, eval = FALSE}
highest <- read_html("https://en.wikipedia.org/wiki/List_of_highest-grossing_films") %>% 
  html_table()
```

## Nested Structures

#### JSON

Web-based graphics started getting popular not too long ago. Generally, stats people were not using them, but web developer-type folks were. They needed a structure that would work well for the web and interact with their JavaScript-based graphics -- thus, JavaScript Object Notation (JSON) was born. You will see JSON come out of many web-based interfaces. 

<a href="https://www3.nd.edu/~sberry5/data/optionsDataBrief.json">This</a> is what JSON looks like.

There are a few JSON-reading packages in R, but jsonlite tends to work pretty well.

```{r, eval = FALSE}
fjsonTest <- jsonlite::read_json(path = "https://www3.nd.edu/~sberry5/data/optionsDataBrief.json", 
                                simplifyVector = TRUE)
```

This is a very simple form of JSON.

#### JSON Dangers

There is JSON and then there is JSON. You might find yourself some interesting data and want to bring it in, but an error happens and you have no idea why the read_json function is telling you that the file is not JSON. 

Not all JSON is pure JSON! When that is the case, you will need to create pure JSON.

Look at this file: https://www3.nd.edu/~sberry5/data/reviews_Musical_Instruments_5.json

It looks like JSON, but...

```{r, eval = FALSE}
jsonlite::validate("https://www3.nd.edu/~sberry5/data/reviews_Musical_Instruments_5.json")
```

If we would want to read that in as true JSON, we would need to do some work:

```{r, eval = FALSE}
musicalInstruments <- readLines("https://www3.nd.edu/~sberry5/data/reviews_Musical_Instruments_5.json")

musicalInstruments <- paste(unlist(lapply(musicalInstruments, function(x) {
  paste(x, ",", sep = "")
})), collapse = "")

musicalInstruments <- paste("[", musicalInstruments, "]", sep = "")

musicalInstruments <- gsub("},]", "}]", musicalInstruments)
```

## Mass Reading

Everything we just learned is great and you will use them all in your data wrangling missions.

Fortunately (or unfortunately, depending on how you look at it), it is not the whole story -- you will frequently be reading in many files of the same time.

If you have two files, you might be able to get away with brute force:

```{r, eval = FALSE}
# DO NOT RUN:

myData1 <- read.csv("test.csv")

myData2 <- read.csv("test2.csv")
```

Would you want to do this for 5 files? What about 100? Or 1000? I will answer it for you: no!

The chunks below introduce some very important functions. We are going to see <span class="func">lapply</span> again -- it is important that you learn to love the apply family!

```{r, eval = FALSE}
# DO NOT RUN:

allFiles <- list.files(path = "", all.files = TRUE, full.names = TRUE, 
                      recursive = TRUE, include.dirs = FALSE)

allFilesRead <- lapply(allFiles, 
                       function(x) read.csv(x, stringsAsFactors = FALSE))

allData <- do.call("rbind", allFilesRead)
```